\normalsize{A general approach and tool to evaluate the effectiveness of no-take marine reserves}

### 2017-07-23 Manuscript.Rmd
Finalized manuscript

---
title: \normalsize{A general approach and tool to evaluate the effectiveness of no-take marine reserves}
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
output:
  pdf_document:
    fig_caption: yes
fontsize: 12pt
bibliography:
- exported-references.bib
- packages.bib
csl: plos.csl
---

@@

Marine reserves are often implemented to preserve habitat and recover overfished stocks.

To date, the evaluation of the reserves relies solely on biological data, ignoring socioeconomic and governance dimensions. In absence of a standardized methodology, existing data is analyzed in different ways, hindering the ability to compare results across communities. We developed a framework to evaluate the effectiveness of the reserves by matching 7 commonly stated management objectives to a set of 9 biophysical, 5 socioeconomic, and 14 governance indicators. This framework includes a standardized survey to collect governance and socioeconomic data. Biophysical indicators are evaluated with a Difference-in-Difference analysis, estimating the net effect of the reserve through time. Linear regression models are fitted to socioeconomic indicators through time, testing for the difference in trends before and after the implementation of the reserve. Governance indicators are analyzed based on literature, identifying common governance structures and their associated effectiveness. To make the framework accessible to fishers and decision makers, and allow replication of results, we developed an open source application. The results produced by the application contain a report for a general audience, and a technical report that can inform the renewal and decision-making process.

@@

Overfishing and unsustainable fishing practices are some of the major threats to the conservation of marine ecosystems around the world [@pauly_2005-qV; @halpern_2008-dK]. Marine Protected Areas (MPAs) are frequently proposed as fishery management and conservation tools to help fish stocks rebound [@lester_2008-F_; @lester_2009-Ks: @sala_2016-PV] by limitting or restricting fishing effort and gears. No-take marine reserves (marine reserves hereinafter) are a particular type of MPAs, where all fishing effort and extractive activities are off-limits or highly limited. The International Union for the Conservation of Nature [-@iucn_website_2017-Bg] categorizes them as Ia (Strict Natural Reserves) or Ib (Wilderness Area). However, see [@hortaecosta_2016-_3] for a detailed regulation-based classification system.

MPAs have proven to increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience of the bounded region [@micheli_2012-EU], and preserve genetic diversity [@munguavega_2015-yg]. Compared to partially protected MPAs, marine reserves are known to have even higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_]. Often, these effects are meassured as biological changes in the area through time and often lack a control site against which to compare [@betti_2017-lq]. This before-after comparison leaves aside other factors for which one must control [@davies_2017-ml], impeding us to talk about causation with complete certainty. While some studies have used control sites, their analyses do not estimate the net effect of the reserve, and often use a control-impact comparison approach [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya]. A smaller fraction of studies use measurements of some biological atribute of a reserve and control sites before and after the implementation of the reserves, thus having a before-after-control-impact design [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], allowing them to use causal inference techniques to estimate the net effect of the reserve.

The diversity of approaches used to evaluate the effect of protected areas highlights the need to standardize approaches that enable us to truly evaluate the effect of protection. Though biophysical aspects are important to reserve success, the effectiveness of reserves also depends on the socioeconomic status and governance system of the local fishing community [@basurto_2013-oq; @basurto_2012-Mz], dimensions largely ignored until recently. Currently, the only the IUCN framework [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biophysical, socioeconomic, and governance indicators, and insight into how these indicators may be measured. However, it does not expand on how to analyze them, nor provide a user-friendly tool that enable replicability ans scalability of the analysis.

Open science and the development of tools have become ingreasingly popular to make science scalable and reproducible [@lowndes_2017-xh]. The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, has succesfully standardized a way to measure the health and benefits of the oceans. This has been succesfuly implemented at global scales, but also at country-level [@selig_2015-F9], and regional scales [@halpern_2014-lQ; @elfes_2014-RC]. Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance indicators [@anderson_2015-ND], design territorial use rights for fisheries [TURFs; @oyanedel_2017-TO], or to improve decision making in the hydropower industry [@vilela_2017-Zo], among others.

The lack of a comprehensive framework -or the complexity of existing ones, which alienate non-experts- and user-friendly tools to evaluate the effectiveness of marine reserves calls for the development of a new framework and tool. Here, we present a framework to evaluate no-take marine reserves, which incorporates the biological, socioeconomic, and governance dimensions of these conservation areas. We provide a list of commonly stated management objectives, and match them to appropriate indicators that meassure the effectiveness of the management intervention. We include a simple approach to analyze these indicators, building on causal inference techniques [@moland_2013-VP], largely needed to truly understand the effect of management interventions [@burgess_2016-HN]. At the same time, we introduce the Marine Reserve Evaluation App (MAREA), an open source web-based tool that automates the framework described in this paper. Case studies from three regions of Mexico are presented as proof of concept of the app.

@@

Here, we describe the proposed framework to evaluate the effectiveness of marine reserves (**Fig. 1**). We explain how management objectives were identified and matched to appropriate indicators that allow the evaluation of the reserves, and provide guidelines on data collection and formating. Alongside, methodologies to analyze these indicators are presented. Then, we describe the development of MAREA and provide guidance on how to use this user-friendly open access tool. Finally, we provide guidelines on how to interpret the results and output generated by MAREA.

@@

Throughout this work, we will refer to the stated goals for which a reserve has been established as objectives. Given that this work was motivated by the need to provide a framework to evaluate mexican marine reserves, we focused on identifying objectives of Mexican marine reserves. However, we group these objectives into seven major categories, which can be applied to other conservation areas in the world. The list of objectives was gathered through a literature review from the reserves’ stated objectives in official documents such as the Technical Justification Studies (*Estudios Tecnicos Justificativos*), agreements, decrees, and specific legislation [@nom049sagpesc_2014-V6]. Even though each reserve -or set of reserves- has its own goals, seven main categories of objectives were identified:

@@

Indicators are divided into three main types: biophysical, socioeconomic, and governance. Biophysical indicators (n = 7) focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data section for specific sampling design and methodologies). Socioeconomic indicators (n = 3) reflect the performance of the fishery in terms of catches, income from catches and availability of alternative livelihoods. Governance indicators (n = 15) describe the governance structures under which the community operates. These indicators may be numeric (*e.g.* Fish biomass) or descriptive (*e.g.* Reasoning for reserve location). Our list includes indicators that respond to the implementation of the reserve (*i.e.* outcome variables) or that might further the understanding of its performance. In that sense, most biophysical and socioeconomic indicators are outcome variables, while governance indicators might explain poor reserve performance, for example. Whenever an indicator is said to also be applied to “Objective Species”, it means that the indicator will be used for all species (*e.g.* Fish Biomass) and individual species that are either the conservation target of the reserve, or are of particular economic or ecological interest (*e.g.* Grouper Biomass). **Table 1** presents the proposed indicators. **Table 2** shows how objectives are matched with indicators.

@@

Biophysical data is collected via underwater ecological surveys as part of a reserve's monitoring program, often carried out by local fishers with guidance of non-government organizations, such as Comunidad y Biodiversidad or Niparajá. Local fishers,trained in scientific diving, perform SCUBA dives to record fish and invertebrate richness and abundances, as well as fish total length. This information is recorded along 30 m transects, with a sampling window of 2 m X 2 m. Between 10 and 24 transects are performed in each site at depths between 1 and 30 m. Ecological surveys are performed yearly in each reserve and its corresponding control site(s), before and after the implementation of the reserve. Control sites are areas where habitat is similar to that of the reserve, but with presence of fishing activity.

Having control sites for biophysical data allows us to use counterfactual approaches to evaluate the net effect of the reserve. The hypothesis that the indicators will respond the the implementation of the reserve is tested by analysing spatial and temporal changes in each numeric biophysical indicator (all but B5) using generalized linear models [@moland_2013-VP]. To account for variations in the environment and survey conditions, covariates that describe the environment and gathered during the underwater ecological surveys are included into a model with form:

@@

In this model, $i$, $t$, and $z$ are subindices for transect, time, and zone (control or reserve site), respectively. This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), tratment ($Z$), an interaction between a variable that indicates pre- or post-implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$), visibility during the survey ($V$), and depth at which survey was performed ($D$). Here, years are modeled as factors, using the first year as the reference level. The treatment and post variables are modeled by dummy variables coded as Control = 0 and Reserve = 1; and Pre-implementation = 0 and Post-implementation = 1, respectively.

Socioeconomic data is collected by the fishers or fishery management agencies, which record landings, income, and sometimes prices for each species. In order to control for inflation and changes in buying power, income is adjusted with the countries consumer price index:

@@

Where $I_t$ represents the adjusted income for year $t$ as the product between the reported income for year $t$ and the ratio between the consumer price index ($CPI$) in that year to the present year's ($T$) CPI. Since no control sites are available, numeric socioeconomic indicators (G1 and G2) are evaluated with a simplified version of *eq. 1*:

@@

While this model does not allow assertion of a causal relation, we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input.

Governance data, however, is not readily available nor systematically gathered by the community or any other organizations. In this case, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1-G15). The survey is included as supplementary material (**Appendix 1**). To analyze governance information, we performed a literature review of common governance structures and ther relation to effectiveness in managing fisheries or marine reserves. These approach has prooven to succesfuly allow the evaluation of governance structures [@espinosaromero_2014-PY]. Governance information is not analyzed in a quantitative way, but is presented along with the biophysical and socioeconomic indicators to provide managers and users with a full description of the reserve.

@@

MAREA was developed in R Studio [@rcore_2017] using the shiny framework [@shiny_2017]. This provides the tools to build interactive web applications, hosted on an open server. MAREA can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/). While the original version was developed in spanish (aimed mainly at Latinamerican countries), all its content can be translated by a translation tool availabe within the environment.

MAREA is designed into a 6-step process. The first part provides an introduction to the app, and provides the user with an overview of the process. Then, the user must select management objectives, and MAREA will automatically select appropriate indicators, based on **Table 2**. Then, the user can load the data, using standard \*.csv formatting; sample datasets are provided in the environment. Once data have been loaded, MAREA identifies all reserves in the data, and lets the user select the reserve to be evaluated. At this point, the user can also specify the year of implementation of the reserve, as well as select objective species that are of particular management interest. Before presenting the results MAREA provides the user with a section to confirm that all the decisions taken before are correct. Finally, the user is taken to the results, which are represented in two ways.

The first result is a color-coded scorecard, intended to provide a general overview of the effectiveness of the reserve to a general public. The scorecard provides a global score for the reserve, a general score for each category of indicators, and an individual score for each selected indicator. The global and category scores are determined by the percentage of positive indicators, overall and for each category. For numeric biophysical indicators (all but B5), the color is defined by the sign of the interaction term coefficient ($\beta_2$) in *eq. 1*. For socioeconomic indicators, colors are assigned based on the direction of the slope ($\beta_1$). Red, yellow or green are used for $\beta_i<0$, $\beta_i = 0$, and $\beta_i>0$ respectively. The intensity of the color is defined by the significance of coefficient, testing the null hypothesis of no change (*i.e.* $H_0: \beta_i = 0$) with Student t-test. Cutoff values are $p < 0.05$ and $p < 0.1$. Thus, even in a case where $\beta_i > 0$ if the coefficient is not significant (*i.e. * $p>0.1$) the indicator will be assigned a yellow color. A legend (**Fig. 2**) is provided in the app to aid in the interpretation of these results. Governance indicators, however, are evaluated binomially and can be either red or green. The color is defined based on what literature shows to be a positive (green) or negative (red) value for each indicator. However, due to the nature of some governance indicators, which require the user to provide a narrative, only indicators XXX are presented in the scorecard.

@@

The second output from MAREA is a technical report, intended to communicate information and statystical results in a rigurous way. This report also includes a scorecard as a summary of the results, but provides more information for each indicator. For all numeric indicators, the technical report includes graphs of the value of the indicator, for the reserve and control sites through time. It also provides a regression table [@stargazer_2015] that summarizes the value of all coefficients in the regression and their respective standard errors. The summary table also provides information on model fit ($R^2$) and significance of the regression.

@@

We apply the proposed framework and tool to evaluate the effectiveness of one no-take marine reserves from Isla Natividad, in Baja California Sur, Mexico. Isla Natividad is located 8 Km off the Pacific Coast of the Baja Peninsula (**Fig. 3**). Fishers in Isla Natividad operate under a fishing cooperative (S.C.P.P. Buzos y Pescadores de la Baja California), which promotes co-management of marine resources [@mccay_2017-1m; @mccay_2014-CN]. Additionaly, fishers have Territorial Use Rights for Fisheries (TURFs), that provide them with exclusive rights to exploit the marine resources for which they hold permits.

@@

In 2006, the community implemented two marine reserves within their TURF. These reserves have prooven to be effective in enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of highly valuable commercial species, like abalone [@munguavega_2015-yg]. These ecological benefits have been translated into economical benefits, enhancing population persistance and supporting abalone fisheries [@rossetto_2015-V0]. For the purpose of this evaluation, we will focus on the "La Plana / Las Cuevas" Marine reserve, located in the southern end of the Island (**Fig. 3**) and its corresponding control site "La Dulce / Babencho".

The reserve was implemented to recover species of economic interest -which have been overexplotied in the last years- in order to enhance fishery production in nearby waters. Fishers were also interested in preserving biological diversity and the ecosystem. Thus, objectives objectives 4-7 were selected. Using **Table 2** to match these objectives with appropriate management indicators, we selected all biophysical, socioeconomic, and governance indicators.

@@

All invertebrates show increased densities within the reserve, as compared to the control site, providing evidence that the reserve is effectively protecting these species.  While we are not able to infer causality from the socioeconomic analysis, we are able to identify important trends. For example, snail and urchin density hve significantly increased within the reserve, but their landings and income have decreased. The opposite is observed for lobster and cucumber, which have shown increases in densities, landing, and income. While further information on market behaviour of each species is needed, these results provide insights into the state of the reserve, as well as the associated fisheries.

@@

The proposed methodologies, specially the way in which biophysical indicators are evaluated, provide valuable information for managers. The analysis isolates the net effect of the reserve, providing a propper measure of reserve effectiveness. We agree there is room for improvement in the way in which socioeconomic and governance data are analized. However, we believe that providing a unifying platform where all can be analyzed and comprehensively presented represents a valuable step towards effective management.

The main value of MAREA is that it provides a free, simple, and replicable way to perform rigurous analysis. The tool can easily be used by fishers, NGO members, and managers in government agencies, providing transparency of the analysis and results. The way in which results are presented allow this information to be interpreted by a wider audience. The scorecard is easily understandable by experts and non-experts, and can be used as an effective tool for communicating the results of yearly evaluations. On the other side, the technical report can serve as a tool for managers and scientists to rapidly communicate information at a more technical level.

>>>>

---
title: \normalsize{A general approach and tool to evaluate the effectiveness of no-take marine reserves}
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
output:
  pdf_document:
    fig_caption: yes
fontsize: 12pt
bibliography:
- exported-references.bib
- packages.bib
# csl: plos.csl
---

@@

Marine reserves are often implemented to preserve habitat, recover overfished stocks, and secure livelihoods of coastal communities. As with any other management interventions, we need to better understand their effectiveness and impacts on coastal communities and the environment. To date, the evaluation of the reserves largely relies on analysing ecological data, ignoring socioeconomic and governance dimensions. In absence of a standardized methodology, existing data is analyzed in different ways, hindering the ability to compare results across case studies. We present a framework and tool to evaluate the effectiveness of marine reserves by matching 7 commonly stated management objectives to a set of 9 biophysical, 5 socioeconomic, and 14 governance indicators. We provide guidelines on how to properly colet data that can then be analysied with standardized methods Biophysical indicators are evaluated with causal inference techniques, using a counterfactual approach, to measure the net effect of the reserve on each indicator. Linear regression models are fitted to socioeconomic indicators through time, testing for the differences before and after the implementation of the reserve. Governance indicators are quantitatively analyzed based on literature, identifying common governance structures and their associated effectiveness. To make the framework accessible to fishers and decision makers, and allow replication of results, we developed an open source, web-based Marine Reserve Evaluation App (MAREA). Together, this framework and MAREA represent a major step in marine reserve evaluation that can furhter our understanding and simplify our management of marine reserves.

@@

Overfishing and unsustainable fishing practices are some of the major threats to the conservation of marine ecosystems around the world [@pauly_2005-qV; @halpern_2008-dK]. Marine Protected Areas (MPAs) are frequently proposed as fishery management and conservation tools to help fish stocks rebound [@lester_2008-F_; @lester_2009-Ks: @sala_2016-PV] by limitting or restricting fishing effort and gears. No-take marine reserves (marine reserves hereinafter) are a particular type of MPAs, where all fishing effort and extractive activities are off-limits or highly regulated. The International Union for the Conservation of Nature [-@iucn_website_2017-Bg] categorizes them as Ia (Strict Natural Reserves) or Ib (Wilderness Areas). However, see [@hortaecosta_2016-_3] for a detailed regulation-based classification system.

MPAs have proven to increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience of the bounded region [@micheli_2012-EU], and preserve genetic diversity [@munguavega_2015-yg]. Compared to partially protected MPAs, marine reserves are known to have even higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_]. Often, these effects are meassured as biological changes in the area through time and lack a control site against which to compare [@betti_2017-lq]. This before-after comparison ignores other factors for which one must control [@davies_2017-ml], impeding us to talk about causation with complete certainty. While some studies have used control sites, their analyses do not estimate the net effect of the reserve, and often use a control-impact comparison approach [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD]. A smaller fraction of studies use measurements of some biological atribute of a reserve and control sites before and after the implementation of the reserves, thus having a before-after-control-impact design [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], allowing them to use causal inference techniques that estimate the net effect of the reserve.

As with any other policy intervention, it is important that we can measure its effect in order to adapt and learn. The diversity of approaches currently used to evaluate the effect of marine reserve often does not answer the simplest question: What is the net effect of the marine reserve? This highlights the need to develop standardized approaches that enable us to truly evaluate the effect of the intervention. Furthermore, while biophysical aspects are important to reserve success, the effectiveness of reserves also depends on the socioeconomic status and governance system of the local fishing community [@basurto_2013-oq; @basurto_2012-Mz], dimensions often ignored. By excluding these important dimensions. the evaluation provides just a partial picture of the state of the reserve. Currently, only the IUCN framework "How to evaluate your MPA" [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biophysical, socioeconomic, and governance indicators, and insight into how these indicators may be measured, but does not provide guidelines on how to analyse them. A more recent work by @mascia_2017-m_ integrates these three dimensions and uses causal inference techniques to provide a measure of the effect of implemented an MPA. However, these two certainly novel approaches do not provide a user-friendly tool that enable replicability ans scalability of the analysis.

An increasingly popular way to mace science reproducible, scalable and replicable is through Open science and the development of tools [@lowndes_2017-xh]. The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, has succesfully standardized a way to measure the health and benefits of the oceans. This has been implemented at global scales, but also at country-level [@selig_2015-F9], and regional scales [@halpern_2014-lQ; @elfes_2014-RC]. Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance indicators [@anderson_2015-ND], design territorial use rights for fisheries [TURFs; @oyanedel_2017-TO], or to improve decision making in the hydropower industry [@vilela_2017-Zo], just to list a few.

The lack of a comprehensive framework -or the complexity of existing ones, which alienate non-experts- and user-friendly tools to evaluate the effectiveness of marine reserves calls for the development of a new framework and tool. Here, we present a framework to evaluate no-take marine reserves, which incorporates the biological, socioeconomic, and governance dimensions of these conservation areas. We provide a list of commonly stated management objectives, and match them to appropriate indicators that meassure the effectiveness of the management intervention. We include a simple approach to analyze these indicators, building on causal inference techniques [@moland_2013-VP], largely needed to truly understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_]. We also introduce the Marine Reserve Evaluation App (MAREA), an open source web-based tool that automates the framework described in this paper. To demonstrate the potential of MAREA, we perform an evaluation of marine reserve established by fishers from Isla Natividad, Mexico in 2006.

@@

Here, we describe the proposed framework to evaluate the effectiveness of marine reserves (**Fig. 1**). We explain how management objectives were identified and matched to appropriate indicators that allow the evaluation of the reserves, and provide brief guidelines on data collection. Alongside, methodologies to analyze these indicators are presented. Then, we describe the development of MAREA and eplain how this user-friendly open access tool can be used by anyone. Finally, we provide guidelines on how to interpret the results and output generated by MAREA.

@@

Throughout this work, we will refer to the stated goals for which a reserve was designed as objectives. Given that this work was motivated by the need to provide a framework to evaluate Mexican marine reserves, or focus was on identifying objectives of marine reserves in Mexico. However, we group these objectives into seven major categories, which can be applied to other conservation areas in the world. The list of objectives was gathered through a literature review from the reserves’ stated objectives in official documents such as the Technical Justification Studies (*Estudios Tecnicos Justificativos*), agreements, decrees, and specific legislation [@nom049sagpesc_2014-V6] associated to these areas. Even though each reserve has its own goals, seven main categories of objectives were identified:

@@

Indicators are divided into three main categories: biophysical, socioeconomic, and governance. Biophysical indicators (n = 7) focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data section for specific sampling design and methodologies). Socioeconomic indicators (n = 3) reflect the performance of the fishery in terms of catches, income from catches, and availability of alternative livelihoods. Governance indicators (n = 15) describe the governance structures under which the community operates. These indicators may be numeric (*e.g.* Fish biomass) or descriptive (*e.g.* Reasoning for reserve location). Our list includes indicators that respond to the implementation of the reserve (*i.e.* outcome variables) or that might further the understanding of its performance. In that sense, most biophysical and socioeconomic indicators are outcome variables. On the other hand, governance indicators are viewed as explanatory variables that might explain reserve performance. Whenever an indicator is said to also be applied to “Objective species”, it means that the indicator can used for all species (*e.g.* Fish Biomass) and individual species that are either the conservation target of the reserve or are of particular economic or ecological interest (*e.g.* Grouper Biomass). **Table 1** presents the proposed indicators, and **Table 2** shows how objectives are matched with indicators.

@@

Biophysical data is collected via underwater ecological surveys as part of a reserve's monitoring program, often carried out by local fishers with guidance of non-government organizations, such as Comunidad y Biodiversidad or Niparajá. Local fishers -trained in scientific diving- perform SCUBA dives to record fish and invertebrate richness and abundances, as well as fish total length. This information is recorded along 30 m transects, with a sampling window of 2 m X 2 m. Between 10 and 24 transects are performed in each site at depths between 1 and 30 m. Ecological surveys are performed yearly in each reserve and its corresponding control site(s), before and after the implementation of the reserve, allowing us to have a before-after-control-impact sampling design. Control sites are areas where habitat is similar to that of the reserve, but with presence of fishing activity.

Having control sites for biophysical data allows us to use counterfactual approaches to evaluate the net effect of the reserve. The hypothesis that the indicators will respond the the implementation of the reserve is tested by analysing spatial and temporal changes in each numeric biophysical indicator (all but B5) using generalized linear models [@moland_2013-VP]. To account for variations in the environment and survey conditions, covariates that are gathered during the underwater ecological surveys are included into a model with form:

@@

In this model, $i$, $t$, and $z$ are subindices for transect, time, and zone (control or reserve site), respectively. This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), a dummy variable that indicates tratment ($Z$; *i.e.* control o reserve), an interaction between a dummy variable that indicates pre- or post-implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$; in °C), horizontal visibility during the survey ($V$; in m), and depth at which survey was performed ($D$; in m). $\epsilon$ represents the error term associated to the equation. Here, years are modeled as factors, using the first year as the reference level. By modelling years as factors, we avoid overimposing a linear structure in the way an indicator changes through time (*i.e.* the change in biomass between 2000 and 2001 does not have to be the same as the change between 2006 an 2007). The treatment and post variables, modeled as dummy variables, are coded as Control = 0 and Reserve = 1; and Pre-implementation = 0 and Post-implementation = 1, respectively.

Socioeconomic data is often collected by fishers or fishery management agencies, which record landings, income, and sometimes prices for each species. In order to control for inflation and changes in buying power, income is adjusted with the countries consumer price index:

@@

Where $I_t$ represents the adjusted income for year $t$ as the product between the reported income for that year and the ratio between the consumer price index ($CPI$) in that year to the most recent year's ($T$) CPI. Since no control sites are available, numeric socioeconomic indicators (G1 and G2) are evaluated with a simplified version of *eq. 1*:

@@

While this model does not allow assertion of a causal relation, we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input. For both models (**eq. 1** and **eq. 3**), coefficients are estimated via ordinary least squares, and heteroskedastik-robust standard errors are calculated.

Governance data, however, is not readily available nor systematically collected by the community or any other organizations. In this case, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1-G15). The survey is included as supplementary material (**Appendix 1**). To analyze governance information, we performed a literature review of common governance structures and ther relation to effectiveness in managing fisheries or marine reserves. These approach has prooven to succesfuly allow the evaluation of governance structures [@espinosaromero_2014-PY]. Governance information is not analyzed in a quantitative way, but is presented along with the biophysical and socioeconomic indicators to provide managers and users with a full description of the reserve.

@@

MAREA was developed in R Studio [@rcore_2017] using the shiny framework [@shiny_2017], which provides the tools to build interactive web applications, hosted on an open server. MAREA can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/). While the original version was developed in spanish (aimed at Mexico and other Latinamerican countries), all its content can be translated by a translation widget availabe within the environment.

MAREA is designed into a 6-step process, divided in tabs. The first tab provides an introduction to the app and an overview of the evaluation process. Then, the user selects management objectives, which MAREA automatically matches to appropriate indicators, based on **Table 2**. The user can then load the data, using standard \*.csv files; sample datasets are provided within MAREA. Once data have been loaded, MAREA identifies all reserves in the data, and lets the user select the reserve to be evaluated. At this point, the user can also specify the year of implementation of the reserve, and indicate objective species that are of particular management interest. Before presenting the results, MAREA provides the user with a section to confirm that all the decisions taken until then are correct. Finally, the user is taken to the results tab, where results can be viewed at a glance or downloaded as a technical report.

The first result is a color-coded scorecard, intended to provide a general overview of the effectiveness of the reserve to a general public. The scorecard provides a global score for the reserve, a general score for each category of indicators, and an individual score for each indicator. The global and category-level scores are determined by the percentage of positive indicators, overall and for each category. For numeric biophysical indicators (all but B5), the color is defined by the sign of the interaction term coefficient ($\beta_2$) in *eq. 1*. For socioeconomic indicators, colors are assigned based on the direction of the slope ($\beta_1$). Red, yellow and green are used for $\beta_i<0$, $\beta_i = 0$, and $\beta_i>0$, respectively. The intensity of the color is defined by the significance of coefficient, testing the null hypothesis of no change (*i.e.* $H_0: \beta_i = 0$) with a Students t-test. Cutoff values are $p < 0.05$ and $p < 0.1$. Thus, even in a case where $\beta_i > 0$ if the coefficient is not significant (*i.e. * $p>0.1$) the indicator will be assigned a yellow color. A legend (**Fig. 2**) is provided within the scorecard to aid in the interpretation of these results. Governance indicators, however, are represented by red or green. The color is defined based on what literature shows to be a positive (green) or negative (red) factor for a reserve. However, due to the nature of some governance indicators, which require the user to provide a narrative, only some indicators are presented in the scorecard (although all are included in the techniical report).

@@

The second output from MAREA is a technical report, intended to communicate information and statystical results in a more formal and technical way. This report also includes a scorecard as a summary of the results, but provides more information for each indicator. For all numeric indicators, the report includes a graph of the value of the indicator, for the reserve and control sites, through time. It also provides a regression table that summarizes the value of all coefficients in the regression and their respective robust standard errors. The summary table also provides information on model fit ($R^2$) and significance of the regression.

@@

We apply the proposed framework and tool to evaluate the effectiveness of one marine reserve from Isla Natividad, in Baja California Sur, Mexico. Isla Natividad is located 8 Km off the Pacific Coast of the Baja Peninsula (**Fig. 3**), where fishers operate under a fishing cooperative (S.C.P.P. Buzos y Pescadores de la Baja California) that promotes co-management of marine resources [@mccay_2017-1m; @mccay_2014-CN]. Additionaly, fishers have Territorial Use Rights for Fisheries (TURFs) that provide them with exclusive access rights to exploit the marine resources in a given perimeter.

@@

In 2006, the community implemented two marine reserves within their TURF [making them TURF-reserves; @afflerbach_2014-HP]. The reserves from Isla Natividad have prooven to be effective in enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of highly valuable commercial species, like abalone [@munguavega_2015-yg]. These ecological benefits have been translated into economical benefits, enhancing population persistance and supporting abalone fisheries [@rossetto_2015-V0]. For the purpose of this evaluation, we will focus on the "La Plana / Las Cuevas" Marine reserve, located in the southern end of the Island (**Fig. 3**) and its corresponding control site "La Dulce / Babencho".

The reserve was implemented to recover species of economic interest -which were overexplotied- in order to enhance fishery production in nearby waters. Fishers were also interested in preserving biological diversity and the ecosystem. Thus, objectives objectives 4-7 were selected. Using **Table 2** to match these objectives with appropriate management indicators, we selected all biophysical, socioeconomic, and governance indicators.

@@

The results identified by our framework and MAREA are consistent to those previously published [*e.g.* @micheli_2012-EU; @rossetto_2015-V0]. Invertebrates show increased densities within the reserve, as compared to the control site, providing evidence that the reserve is effectively protecting these species.  While we are not able to infer causality from the socioeconomic analysis, we are able to identify important trends. For example, snail and urchin density have significantly increased within the reserve, but their landings and income have decreased. The opposite is observed for lobster and cucumber, which have shown increases in densities, landing, and income. While further information on market behaviour of each species is needed, these results provide insights into the state of the reserve, as well as the associated fisheries.

@@

The proposed methodologies, specially the way in which biophysical indicators are evaluated, provide valuable information for managers. The analysis isolates the net effect of the reserve, providing a propper measure of reserve effectiveness. We acknowledge there is room for improvement in the way in which socioeconomic and governance data are analized. Despire this, we believe that providing a unifying platform where all can be analyzed and comprehensively presented represents a valuable step towards evidence-based effective management.

Furthermore, MAREA's value is that it provides a free, simple, and replicable way to perform rigurous analysis. The tool can easily be used by fishers, NGO members, and managers in government agencies, providing transparency of the analysis and results. The way in which results are presented allow this information to be interpreted by a wider audience. The scorecard is easily understandable by experts and non-experts, and can be used as an effective tool for communicating the results of yearly evaluations. On the other side, the technical report can serve as a tool for managers and scientists to rapidly communicate information at a more technical level.
